---
title: "IS606 - Collaborative Project 1"
author: "Mauricio Alarcon, Nathan Lim, Mohamed Elmoudni"
date: "March 3, 2015"
output: html_document
---

* * *

##Table of Contents
* [Background](#background)
* [Overview of Datasets](#dataset_overview)
* [Supply vs Demand](#supply_demand)
* [Seasonal Variation](#seasonal_variation)
* [Increasing James' Profit](#improving_profit)
* [Current and Expected Profit](#expected_profit)
* [Conclusion](#conclusion)

* * *

```{r, echo=FALSE,results = 'hide', warning=FALSE, message=FALSE}
library("ggplot2")
library("boot")
library("pastecs")
library("reshape")
library("lattice")
library("Hmisc")
library("RCurl")

# get files from CSV in github
pricing_csv <- getURL("https://raw.githubusercontent.com/rmalarc/is606_col1/master/details.csv")
pricing <- read.csv(text=pricing_csv,head=TRUE,sep=",",as.is=TRUE)

historical_csv <- getURL("https://raw.githubusercontent.com/rmalarc/is606_col1/master/sales.csv")
historical_data <- read.csv(text=historical_csv,head=TRUE,sep=",",as.is=TRUE)

#Capitalize sandwich type
pricing$type <- as.factor(capitalize(pricing$type))

# properly format date
historical_data$date_dt <- as.POSIXct(historical_data$date, format="%Y-%m-%d")
historical_data$month <- factor(format(historical_data$date_dt, "%b")
                                ,levels=c("Mar","Apr","May","Jun","Jul","Aug"))

# get month, day_of_week and month_day_of_week to see if there is any
# type of variation by any of these variables
historical_data$day_of_week <- factor(weekdays(historical_data$date_dt)
                                      ,levels=c("Monday"
                                                ,"Tuesday"
                                                ,"Wednesday"
                                                ,"Thursday"
                                                ,"Friday")
                                      )
historical_data$month_day_of_week <- paste(historical_data$month, "-" ,historical_data$day_of_week ) 


# pivot the data so we can use it better in GGPLOT
historical_pivoted<-data.frame(date=historical_data$date_dt
                               ,month=historical_data$month
                               ,day_of_week=historical_data$day_of_week
                               ,sandwich="Ham"
                              ,demand=historical_data$demand.ham
                              ,supply=historical_data$available.ham
                              )
historical_pivoted<-rbind(historical_pivoted
                          ,data.frame(date=historical_data$date_dt
                                      ,month=historical_data$month
                                      ,day_of_week=historical_data$day_of_week
                                      ,sandwich="Turkey"
                                      ,demand=historical_data$demand.turkey
                                      ,supply=historical_data$available.turkey)
                          )
historical_pivoted<-rbind(historical_pivoted
                          ,data.frame(date=historical_data$date_dt
                                      ,month=historical_data$month
                                      ,day_of_week=historical_data$day_of_week
                                      ,sandwich="Veggie"
                                      ,demand=historical_data$demand.veggie
                                      ,supply=historical_data$available.veggie)
                          )

```


##<a name="background"></a>Background

James has been operating a sandwich stand in the lobby of his office building during the lunch hour for the past two years.He has been tracking sandwich demand over the two years and he has carefully recorded the number of each type of sandwich demanded, the number he brought with him to sell, and his prices for each type of sandwich.

James provided his data set for analysis.

* * *

##<a name="dataset_overview"></a>Overview of Datasets

###sales.csv
The file `sales.csv` contains the demand and availability by date for each of the three sandwiches that James sells.

```{r}

summary(historical_data[,c("date_dt")])

stat.desc(historical_data[,c("demand.ham","demand.turkey","demand.veggie","available.ham","available.turkey","available.veggie")])

```

As we can see from the data above, mean-demands for each sandwich type are 15.94(Ham), 22.05(Turkey) and 13.06(Veggie). The best selling sandwich is Turkey

###details.csv
This file contains cost and sale prices according to sandwich type:
```{r}
(pricing)
```

* * *

##<a name="supply_demand"></a>Supply vs Demand

From the plot below we can see that the productions of Turkey and Veggie sandwiches are below the demand.

```{r, echo=FALSE}

p <- qplot(sandwich,demand, data=historical_pivoted, geom=c("boxplot", "jitter"),
      fill=sandwich,
      main="Average Supply and Demand by Sandwich Type",
      xlab= "Sandwich Type",
      ylab="Demand")
p + geom_hline(yintercept = c(mean(historical_data$available.ham)
                              ,mean(historical_data$available.turkey)
                              ,mean(historical_data$available.veggie)
                              )
               ,color=c("red","darkgreen","blue")
               ,size=1
               ,linetype=11
               ,label = c("Demand Ham", "Demand Turkey", "Demand Veggie")
               ) +
  scale_fill_discrete(name="Sandwich")

```

```{r, echo=FALSE}
# The total demand and supply
demand_totals <- data.frame(aggregate(demand ~ sandwich
                                  , data = historical_pivoted
                                  , FUN = "sum"
                                  )
                        ,"supply_demand"="Demand"
                        )
demand_totals<-rename(demand_totals,c("demand"="volume"))

supply_totals <- data.frame(aggregate(supply ~ sandwich
                                  , data = historical_pivoted
                                  , FUN = "sum"
                                  )
                        ,"supply_demand"="Supply"
                        )
supply_totals<-rename(supply_totals,c("supply"="volume"))

totals <- rbind(demand_totals,supply_totals)

ggplot(totals
       , aes(factor(sandwich), volume, fill = supply_demand
             )
       ) + 
  geom_bar(stat="identity", position = "dodge",title="") + 
  scale_fill_brewer(palette = "Set2") +
  ggtitle("Total Supply and Demand by Sandwich Type") +
  xlab( "Sandwich") +
  ylab("Volume") +
  theme(legend.title=element_blank())

```

* * *

##<a name="seasonal_variation"></a>Seasonal Variation of Demand

Before we proceed any further, we want to know if there are any seasonal statistically significant variations. We want to prove the significance of any of the following variations by sandwich type:

* From day of the week to day of the week (inter-day - sandwich:day_of_week)
* From month to month (inter-month - sandwich:month)
* From month and day of the week to month and day of the week (inter-month-day - sandwich:month:day_of_week).

We conducted an analysis of variance of the demand in order to accept or reject each of our hypothesis. 

The results are as follows:

```{r}

summary(aov(demand ~ sandwich:month+sandwich:day_of_week+sandwich:month:day_of_week
            ,data=historical_pivoted
            )
        )
```

As we can see in the results above, the only statistically significant variation(normally p<0.05) in the distribution of the demand is the month.

The chart below illustrate the actual fluctuations of the demand by sandwich type and month:

```{r, echo=FALSE}

qplot(month,demand, data=historical_pivoted, geom=c("boxplot"),
      fill=sandwich,
      main="Average Demand by Sandwich Type and Month",
      xlab= "Month",
      ylab="Demand")+
  scale_fill_discrete(name="Sandwich")

#qplot(demand, data=historical_pivoted[historical_pivoted$sandwich=="Ham",], 
#      geom=c("density"),
#      main="Distribution of Ham Sandwich Demand by Month",
#      xlab="Demand",
#      xlim=c(0,40),
#      ylab="Density",
#      facets=month~sandwich, fill=month)+scale_fill_discrete(name="Month")

#qplot(demand, data=historical_pivoted[historical_pivoted$sandwich=="Turkey",], 
#      geom=c("density"),
#      main="Distribution of Turkey Sandwich Demand by Month",
#      xlab= "Demand",
#      xlim=c(0,40),
#      ylab="Density",
#      facets=month~sandwich, fill=month)+scale_fill_discrete(name="Month")
      
qplot(demand, data=historical_pivoted, 
      geom=c("density"),
      main="Distribution of Sandwich Demand by Month",
      xlab= "Demand",
      xlim=c(0,40),
      ylab="Density",
      facets=month~sandwich, fill=sandwich)+scale_fill_discrete(name="sandwich")

#qplot(day_of_week,demand, data=historical_pivoted, geom=c("boxplot"),
#      fill=sandwich,
#      main="Average Demand by Sandwich Type and Day of Week",
#      xlab= "Day of Week",
#      ylab="Demand")+
#  scale_fill_discrete(name="Sandwich")
```

**Based on these findings, we will recommend monthly sandwich produtions to James.**


* * *

##<a name="improving_profit"></a>Increasing James' Profit

James often produces less sandwiches than the actual demand, particularly turkey and veggie sandwiches. 
We also, found that there exists variation in the demand from month to month. Therefore, adjusting his supply levels according to this will help him maximize his profit.

We recommend that James be adjusting his production levels like the following table:

```{r, echo=FALSE}

#calculate the expected demand (use median)
expected_demand <- data.frame(aggregate(demand ~ sandwich+month
                                  , data = historical_pivoted
                                  , FUN = "median"
                                  )
                        )
expected_demand<-rename(expected_demand,c("demand"="expected_demand"))

#bump it up to the ceiling integer
expected_demand$expected_demand<-ceiling(expected_demand$expected_demand)

expected_demand_output_table <- data.frame(cast(expected_demand
                                            , month~sandwich
                                            , first
                                            ,value="expected_demand"
                                            )
                                       )

(expected_demand_output_table)
```

##<a name="expected_profit"></a>Current and Expected Profit

The table below shows James' current profit and expected profit based on the recommended production levels.

```{r, echo=FALSE}

#join the historical data with the expected demand
historical_pivoted <- merge(historical_pivoted, expected_demand, by=c("sandwich","month"))

#join with pricing info
historical_pivoted <- merge(historical_pivoted, pricing, by.x="sandwich",by.y="type")


# calculate expected sales based on expected demand vs actually sold
historical_pivoted$sales_current <- pmin(historical_pivoted$supply,historical_pivoted$demand)

historical_pivoted$sales_expected <- pmin(historical_pivoted$expected_demand,historical_pivoted$demand)

# calculate profit per day based on current and expected demand
historical_pivoted$profit_current <- historical_pivoted$sales_current * historical_pivoted$price - historical_pivoted$supply*historical_pivoted$cost

historical_pivoted$profit_expected <- historical_pivoted$sales_expected * historical_pivoted$price - historical_pivoted$expected_demand*historical_pivoted$cost

#calculate profit summary totals by sandwich type
profit_summary <- data.frame(aggregate(cbind(profit_current,profit_expected)~sandwich
                                  , data = historical_pivoted
                                  , FUN = "sum"
                                  )
                        )

# calculate grand totals
profit_grand_total <- data.frame(sandwich="TOTAL"
                                 ,profit_current=sum(historical_pivoted$profit_current)
                                 ,profit_expected=sum(historical_pivoted$profit_expected)
                                 )

profit_summary <- rbind(profit_summary,profit_grand_total)

profit_summary$increase <- with(profit_summary, sprintf("%3.1f%%"
                                   ,100*(profit_expected-profit_current)/profit_current))
(profit_summary)
```

The expected profit was calculated by applying the recommended production levels to the provided historical data. As we can see, James' profit would have increased by **8.4%**.

* * *

##<a name="conclusion"></a>Conclusion

This recommendation is based on the data that James collected from Mar.2013 to Aug.2013. As we analysed above, the main reason that James could not maximize his profit was on supply deficiency of Turkey and Veggie sandwiches.
Therefore, if he adjust the supply like we recommended, he would make more money, about $180 monthly
