---
title: "IS606 - Collaborative Project 1"
author: "Mauricio Alarcon, Nathan Lim, Simon Elmoudni"
date: "March 3, 2015"
output: html_document
---

* * *

##Table of Contents
* [Background](#background)
* [Overview of Datasets](#dataset_overview)
* [Supply vs Demand](#supply_demand)
* [Seasonal Variation](#seasonal_variation)
* [Improving Profits](#improving_profits)

* * *

```{r, echo=FALSE}
library("ggplot2")
library("boot")
library("plyr")
#library("psych")
library("pastecs")

pricing <- read.csv(file="./details.csv",head=TRUE,sep=",")

historical_data <- read.csv(file="./sales.csv",head=TRUE,sep=",")

# properly format date
historical_data$date_dt <- as.POSIXct(historical_data$date, format="%Y-%m-%d")
historical_data$month <- factor(format(historical_data$date_dt, "%b")
                                ,levels=c("Mar","Apr","May","Jun","Jul","Aug"))

# get month, day_of_week and month_day_of_week to see if there is any
# type of variation by any of these variables
historical_data$day_of_week <- factor(weekdays(historical_data$date_dt)
                                      ,levels=c("Monday"
                                                ,"Tuesday"
                                                ,"Wednesday"
                                                ,"Thursday"
                                                ,"Friday")
                                      )
historical_data$month_day_of_week <- paste(historical_data$month, "-" ,historical_data$day_of_week ) 


# pivot the data so we can use it better in GGPLOT
historical_pivoted<-data.frame(date=historical_data$date_dt
                               ,month=historical_data$month
                               ,day_of_week=historical_data$day_of_week
                               ,sandwich="Ham"
                              ,demand=historical_data$demand.ham
                              ,supply=historical_data$available.ham
                              )
historical_pivoted<-rbind(historical_pivoted
                          ,data.frame(date=historical_data$date_dt
                                      ,month=historical_data$month
                                      ,day_of_week=historical_data$day_of_week
                                      ,sandwich="Turkey"
                                      ,demand=historical_data$demand.turkey
                                      ,supply=historical_data$available.turkey)
                          )
historical_pivoted<-rbind(historical_pivoted
                          ,data.frame(date=historical_data$date_dt
                                      ,month=historical_data$month
                                      ,day_of_week=historical_data$day_of_week
                                      ,sandwich="Veggie"
                                      ,demand=historical_data$demand.veggie
                                      ,supply=historical_data$available.veggie)
                          )

```


##<a name="background"></a>Background

James has been operating a sandwich stand in the lobby of his office building during the lunch hour for the past two years.He has been tracking sandwich demand over the two years and he has carefully recorded the number of each type of sandwich demanded, the number he brought with him to sell, and his prices for each type of sandwich.

James provided his dataset for analysis.

* * *

##<a name="dataset_overview"></a>Overview of Datasets

###sales.csv
The file `sales.csv` contains the demand and availability by date for each of the three sandwiches that James sells.

```{r}

summary(historical_data[,c("date_dt")])

stat.desc(historical_data[,c("demand.ham","demand.turkey","demand.veggie","available.ham","available.turkey","available.veggie")])

```

As we can see from the data above, the data ranges from 2014/03/03 to 2014/08/29(n=130). In addition, the mean demand for each sandwich type ranges from 22.05 (Ham) to 13.06 (Veggie). The best selling sandwich is Ham

###details.csv
The file contains cost and sale pricing by sandwich type as follows:
```{r}
(pricing)
```

* * *

##<a name="supply_demand"></a>Supply vs Demand

From the plot below we can see that the produciton levels for Turkey and Veggie sandwiches are below the demand.

```{r, echo=FALSE}

p <- qplot(sandwich,demand, data=historical_pivoted, geom=c("boxplot", "jitter"),
      fill=sandwich,
      main="Average Supply and Demand by Sandwich Type",
      xlab= "Sandwich Type",
      ylab="Demand")
p + geom_hline(yintercept = c(mean(historical_data$available.ham)
                              ,mean(historical_data$available.turkey)
                              ,mean(historical_data$available.veggie)
                              )
               ,color=c("red","darkgreen","blue")
               ,size=1
               ,linetype=11
               ,label = c("Demand Ham", "Demand Turkey", "Demand Veggie")
               )

```

```{r, echo=FALSE}
# total the demands and supply
demand_totals <- data.frame(aggregate(demand ~ sandwich
                                  , data = historical_pivoted
                                  , FUN = "sum"
                                  )
                        ,"supply_demand"="Demand"
                        )
demand_totals<-rename(demand_totals,c("demand"="volume"))

supply_totals <- data.frame(aggregate(supply ~ sandwich
                                  , data = historical_pivoted
                                  , FUN = "sum"
                                  )
                        ,"supply_demand"="Supply"
                        )
supply_totals<-rename(supply_totals,c("supply"="volume"))

totals <- rbind(demand_totals,supply_totals)


ggplot(totals
        ,color=c("red","darkgreen","blue")
       , aes(factor(sandwich), volume, fill = supply_demand
             )
       ) + 
  geom_bar(stat="identity", position = "dodge",title="") + 
  scale_fill_brewer(palette = "Set2") +
  ggtitle("Total Supply and Demand by Sandwich Type") +
  xlab( "Sandwich") +
  ylab("Volume") +
  theme(legend.title=element_blank())

```

* * *

##<a name="seasonal_variation"></a>Seasonal Variation

As we can see from the chart above, even though there seems to be some variation in the order volume by month.

```{r}

qplot(month,demand, data=historical_pivoted, geom=c("boxplot"),
      fill=sandwich,
      main="Average Demand by Sandwich Type and Month",
      xlab= "Sandwich Type",
      ylab="Demand")
```

### Analysis of Variance of Seasonal Variation
This leads us to see which of these differences are statistically significant. For that matter we then conducted an analysis of variance on the demand by sandwich type and month, day of week and month-day of week. 

The results are as follows:

```{r}

summary(aov(demand ~ sandwich:month+sandwich:day_of_week+sandwich:month:day_of_week
            ,data=historical_pivoted
            )
        )
```

As we can see in the results above, the only statistical variation in the distribution of the demand is attributed to the sandwich type and month.  

**Based on this result we will use monthly volumne by sandwich type to determine the expected value for the recomendation.**

* * *

##<a name="improving_profits"></a>Improving Profits

Based on the review above we recomend that James adjustes his production levels to better meet the actual demand as follows:



